Cardinality estimation methods for HyperLogLog sketches

### 2016-05-23 paper/paper.tex
implemented estimation algorithm that uses precalculated corrections added new algorithm to paper

\symDefine{\symPrecision}{p}
\symDefine{\symRegRange}{q}
\symDefine{\symRegVal}{k}
\symDefine{\symRegValVariate}{K}
\symDefine{\symNumReg}{m}
\symDefine{\symDataItem}{D}
\symDefine{\symBitRepA}{a}
\symDefine{\symBitRepB}{b}
\symDefine{\symIndexI}{i}
\symDefine{\symIndexK}{k}
\symDefine{\symIndexL}{l}
\symDefine{\symIndexJ}{j}
\symDefine{\symCount}{c}
\symDefine{\symAlpha}{\alpha}
\symDefine{\symBeta}{\beta}
\symDefine{\symA}{a}
\symDefine{\symB}{b}
\symDefine{\symE}{e}
\symDefine{\symX}{x}
\symDefine{\symXEstimate}{\hat{\symX}}
\symDefine{\symXNormalized}{\symX'}
\symDefine{\symN}{n}
\symDefine{\symS}{s}
\symDefine{\symY}{y}
\symDefine{\symZ}{z}
\symDefine{\symInteger}{z}
\symDefine{\symIntegerVariate}{Z}
\symDefine{\symCardinality}{n}
\symDefine{\symCardinalityEstimate}{\hat{\symCardinality}}
\symDefine{\symCardinalityOriginalEstimate}{\symCardinalityEstimate_\text{raw}}
\symDefine{\symError}{\varepsilon}
\symDefine{\symStopDelta}{\delta}
\symDefine{\symStopEpsilon}{\varepsilon}
\symDefine{\symBigO}{\mathcal{O}}
\symDefine{\symExpectation}{\mathbb{E}}
\symDefine{\symProbability}{P}
\symDefine{\symProbabilityMass}{\rho}
\symDefine{\symRegProbability}{\gamma}
\symDefine{\symNu}{\nu}
\symDefine{\symKappa}{\kappa}
\symDefine{\symLikelihood}{\mathcal{L}}
\symDefine{\symPoissonRate}{\lambda}
\symDefine{\symPoissonRateEstimate}{\hat{\symPoissonRate}}
\symDefine{\symFunc}{f}
\symDefine{\symFuncPrime}{g}
\symDefine{\symHelper}{h}
\symDefine{\symHelperApprox}{\tilde{\symHelper}}
\symDefine{\symSetA}{A}
\symDefine{\symSetB}{B}
\symDefine{\symSetS}{S}
\symDefine{\symSetX}{X}
\symDefine{\symSetASuffix}{a}
\symDefine{\symSetBSuffix}{b}
\symDefine{\symSetXSuffix}{x}
\symDefine{\symCountMatrix}{\mathbf{c}}
\symDefine{\symPowerSeriesFunc}{\xi}
\symDefine{\symEpsPowerSeriesFunc}{\varepsilon_\symPowerSeriesFunc}

@@

Inserting all these estimators into \eqref{equ:raw_estimate_with_multiplicities} gives
\begin{equation}
\label{equ:correctedestimator}
\symPoissonRateEstimate 
= 
\frac{\symNumReg^2}{2\log(2)\sum_{\symRegVal=-\infty}^\infty \hat{\symCount}'_\symRegVal 2^{-\symRegVal}}
=
\frac{\symNumReg^2}{2\log(2)\left(\symCount_0^{\text{corr}} + \sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal 2^{-\symRegVal} + \symCount_{\symRegRange+1}^{\text{corr}}2^{-(\symRegRange+1)}\right)}
\end{equation}
where $\symCount_0^{\text{corr}}$ and $\symCount_{\symRegRange+1}^{\text{corr}}$ are corrections for $\symCount_0$ and $\symCount_{\symRegRange+1}$, respectively (compare \eqref{equ:intermediate_estimate}). They are given by
\begin{equation}
\symCount_0^{\text{corr}} := \symNumReg
\cdot
\left(
\frac{\symCount_0}{\symNumReg}
+
\frac{1}{2}\sum_{\symRegVal=1}^\infty
\left(\frac{\symCount_0}{\symNumReg}\right)^{2^\symRegVal} 2^{\symRegVal}
\right)
\end{equation}
and
\begin{equation}
\symCount_{\symRegRange+1}^{\text{corr}}
:=
\symNumReg
\cdot
\left(
\sum_{\symRegVal=1}^\infty
\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)	^{2^{-\symRegVal}}
2^{-\symRegVal}
-
\left(
1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}
\right)
\right).
\end{equation}

For the special case of linear probabilistic counting with $\symRegRange=0$, we get for the  estimator \eqref{equ:correctedestimator}
\begin{equation}
\symPoissonRateEstimate 
= 
\frac{\symNumReg}{\log(2)\sum_{\symRegVal=-\infty}^\infty
\left(\frac{\symCount_0}{\symNumReg}\right)^{2^\symRegVal}2^\symRegVal}
=
\frac{\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right)}{\symPowerSeriesFunc\left(\log_2\left(\log\left(\frac{\symNumReg}{\symCount_0}\right)\right)\right)}
\approx
\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right).
\end{equation}
Here we used function $\symPowerSeriesFunc$ defined in \eqref{equ:power_series_function} which is known to be approximately equal to 1. Hence, for this special case we get as expected almost the same estimator as given in \eqref{equ:small_range_estimate}.

>>>>

\symDefine{\symPrecision}{p}
\symDefine{\symRegRange}{q}
\symDefine{\symRegVal}{k}
\symDefine{\symRegValVariate}{K}
\symDefine{\symNumReg}{m}
\symDefine{\symDataItem}{D}
\symDefine{\symBitRepA}{a}
\symDefine{\symBitRepB}{b}
\symDefine{\symIndexI}{i}
\symDefine{\symIndexK}{k}
\symDefine{\symIndexL}{l}
\symDefine{\symIndexJ}{j}
\symDefine{\symCount}{c}
\symDefine{\symAlpha}{\alpha}
\symDefine{\symBeta}{\beta}
\symDefine{\symA}{a}
\symDefine{\symB}{b}
\symDefine{\symE}{e}
\symDefine{\symX}{x}
\symDefine{\symXEstimate}{\hat{\symX}}
\symDefine{\symXNormalized}{\symX'}
\symDefine{\symN}{n}
\symDefine{\symS}{s}
\symDefine{\symY}{y}
\symDefine{\symZ}{z}
\symDefine{\symInteger}{z}
\symDefine{\symIntegerVariate}{Z}
\symDefine{\symCardinality}{n}
\symDefine{\symCardinalityEstimate}{\hat{\symCardinality}}
\symDefine{\symCardinalityOriginalEstimate}{\symCardinalityEstimate_\text{raw}}
\symDefine{\symError}{\varepsilon}
\symDefine{\symStopDelta}{\delta}
\symDefine{\symStopEpsilon}{\varepsilon}
\symDefine{\symBigO}{\mathcal{O}}
\symDefine{\symExpectation}{\mathbb{E}}
\symDefine{\symProbability}{P}
\symDefine{\symProbabilityMass}{\rho}
\symDefine{\symRegProbability}{\gamma}
\symDefine{\symNu}{\nu}
\symDefine{\symKappa}{\kappa}
\symDefine{\symLikelihood}{\mathcal{L}}
\symDefine{\symPoissonRate}{\lambda}
\symDefine{\symPoissonRateEstimate}{\hat{\symPoissonRate}}
\symDefine{\symFunc}{f}
\symDefine{\symFuncPrime}{g}
\symDefine{\symHelper}{h}
\symDefine{\symHelperApprox}{\tilde{\symHelper}}
\symDefine{\symSetA}{A}
\symDefine{\symSetB}{B}
\symDefine{\symSetS}{S}
\symDefine{\symSetX}{X}
\symDefine{\symSetASuffix}{a}
\symDefine{\symSetBSuffix}{b}
\symDefine{\symSetXSuffix}{x}
\symDefine{\symCountMatrix}{\mathbf{c}}
\symDefine{\symPowerSeriesFunc}{\xi}
\symDefine{\symSmallCorrectionFunc}{\sigma}
\symDefine{\symLargeCorrectionFunc}{\tau}
\symDefine{\symEpsPowerSeriesFunc}{\varepsilon_\symPowerSeriesFunc}

@@

Inserting all these estimators into \eqref{equ:raw_estimate_with_multiplicities} gives
\begin{equation}
\label{equ:correctedestimator}
\symPoissonRateEstimate 
= 
\frac{\symNumReg^2}{2\log(2)\sum_{\symRegVal=-\infty}^\infty \hat{\symCount}'_\symRegVal 2^{-\symRegVal}}
=
\frac{\symNumReg^2}
{2\log(2)
\left(
\symSmallCorrectionFunc(\symCount_0) + \sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal 2^{-\symRegVal} + \symLargeCorrectionFunc(\symCount_{\symRegRange+1}) 2^{-(\symRegRange+1)}
\right)}
\end{equation}
where $\symSmallCorrectionFunc(\symCount_0)$ and $\symLargeCorrectionFunc(\symCount_{\symRegRange+1})$ replace $\symCount_0$ and $\symCount_{\symRegRange+1}$ in the \emph{raw} estimate \eqref{equ:intermediate_estimate}, respectively. The functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are given by
\begin{equation}
\symSmallCorrectionFunc(\symCount) := \symNumReg
\cdot
\left(
\frac{\symCount}{\symNumReg}
+
\frac{1}{2}\sum_{\symRegVal=1}^\infty
\left(\frac{\symCount}{\symNumReg}\right)^{2^\symRegVal} 2^{\symRegVal}
\right)
\end{equation}
and
\begin{equation}
\symLargeCorrectionFunc(\symCount)
:=
\symNumReg
\cdot
\left(
\sum_{\symRegVal=1}^\infty
\left(1-\frac{\symCount}{\symNumReg}\right)	^{2^{-\symRegVal}}
2^{-\symRegVal}
-
\left(
1-\frac{\symCount}{\symNumReg}
\right)
\right)
=
\symNumReg
\cdot
\left(
\frac{\symCount}{\symNumReg}
-
\sum_{\symRegVal=1}^\infty
\left(1-\left(1-\frac{\symCount}{\symNumReg}\right)^{2^{-\symRegVal}}\right)
2^{-\symRegVal}
\right)
\end{equation}

For the special case of linear probabilistic counting with $\symRegRange=0$, we get for the  estimator \eqref{equ:correctedestimator}
\begin{equation}
\symPoissonRateEstimate 
= 
\frac{\symNumReg}{\log(2)\sum_{\symRegVal=-\infty}^\infty
\left(\frac{\symCount_0}{\symNumReg}\right)^{2^\symRegVal}2^\symRegVal}
=
\frac{\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right)}{\symPowerSeriesFunc\left(\log_2\left(\log\left(\frac{\symNumReg}{\symCount_0}\right)\right)\right)}
\approx
\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right).
\end{equation}
Here we used function $\symPowerSeriesFunc$ defined in \eqref{equ:power_series_function} which is known to be approximately equal to 1. Hence, for this special case we get as expected almost the same estimator as given in \eqref{equ:small_range_estimate}.
\subsection{Estimation algorithm}
\begin{algorithm}
\caption{Cardinality estimation algorithm based on the corrected \emph{raw} estimate.}
\label{alg:estimate_raw_corrected}
\begin{algorithmic}
\Function {EstimateCardinality}{$\vec{\symCount}$}
\State $\symNumReg \gets \left\Vert\vec{\symCount}\right\Vert_1$
\State $\symZ\gets 0.5\cdot \symLargeCorrectionFunc(\symCount_{\symRegRange+1})$
\For{$\symRegVal\gets \symRegRange, 1$}
\State $\symZ\gets\symZ + 0.5\cdot \symCount_\symRegVal$
\EndFor
\State $\symZ\gets\symZ+\symSmallCorrectionFunc(\symCount_0)$
\State \Return$\symNumReg^2 / \left(2\log(2)\symZ\right)$
\EndFunction
\end{algorithmic}
\end{algorithm}
